\ldots

### 2013-06-03 summer13/frac2.tex
added some background and started working on syntax of pi with fractions

\documentclass{llncs}
\usepackage{url}
\usepackage{proof}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{comment}

\newcommand{\dgm}[2][1.5]{
\begin{center}
\scalebox{#1}{
\includegraphics{diagrams/#2.pdf}
}
\end{center}
}

\newcommand{\todo}[1]{\textbf{TODO:} #1}
\newcommand{\jacques}[1]{\textsc{Jacques says:} #1}
\newcommand{\amr}[1]{\textsc{Amr says:} #1}

\hyphenation{a-reas}

@@

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include Pi
%! [^ = \ulcorner
%! ^] = \urcorner
%! [v = \llcorner
%! v] = \lrcorner
%! [[ = \llbracket
%! ]] = \rrbracket
%! ^^^ = ^{\dagger}
%! eta* = \eta
%! eps* = \epsilon
%! Union = \bigcup
%! in = \in
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |-->let = \mapsto_{let}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! -> = \rightarrow
%! ~> = \leadsto
%! ::= = ::=
%! /= = \neq
%! vi = v_i
%! di = d_i
%! si = s_i
%! sj = s_j
%! F = \texttt{F}
%! T = \texttt{T}
%! forall = \forall
%! exists = \exists
%! empty = \emptyset
%! Sigma = \Sigma
%! eta = \eta
%! where = \textbf{where}
%! epsilon = \varepsilon
%! least = \phi
%! loop+ = loop_{+}
%! loop* = loop_{\times}
%! CatC = {\mathcal C}
%! CatA = {\mathcal A}
%! gamma = \gamma
%! {[ = \{
%! ]} = \}
%! elem = \in
%! dagger = ^\dagger
%! alpha = \alpha
%! beta = \beta
%! rho = \rho
%! @@ = \mu
%! @ = \,@\,
%! Pow = \mathcal{P}
%! Pi = \Pi
%! PiT = \Pi^{o}
%! PiEE = \Pi^{\eta\epsilon}_{+}
%! PiT = \Pi^{o}
%! PiTF = \Pi^{/}
%! bullet = \bullet
%! * = \times

\begin{document}

\section{Higher-Order Reversible Functions}

To get from {{Pi}} types and combinators to a higher order setting, we hope to
add \emph{fractional types} to the language. Recall that to construct the
(positive) rational numbers {{Q}}, we can take pairs of positive integers
{{(m, n)}} and quotient by the equivalence relation that says {{(m1, n1)}} is
equal to {{(m2, n2)}} if {{m1 * n2 = m2 * n1}}. We can write this as an
inference rule as follows:

@@

\subsection{Defining trace}

@@

\subsection{Higher-Order Combinators}

@@

\end{document}

>>>>

\documentclass{sigplanconf}
\usepackage{url}
\usepackage{proof}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{comment}

\newcommand{\dgm}[2][1.5]{
\begin{center}
\scalebox{#1}{
\includegraphics{diagrams/#2.pdf}
}
\end{center}
}
\newcommand{\todo}[1]{\textbf{TODO:} #1}
\newcommand{\jacques}[1]{\textsc{Jacques says:} #1}
\newcommand{\amr}[1]{\textsc{Amr says:} #1}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{proposition}[theorem]{Proposition}

@@

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include Pi
%! [^ = \ulcorner
%! ^] = \urcorner
%! [v = \llcorner
%! v] = \lrcorner
%! [[ = \llbracket
%! ]] = \rrbracket
%! ^^^ = ^{\dagger}
%! eta* = \eta
%! eps* = \epsilon
%! Union = \bigcup
%! in = \in
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |-->let = \mapsto_{let}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! -> = \rightarrow
%! ~> = \leadsto
%! ::= = ::=
%! amp = \&
%! /= = \neq
%! vi = v_i
%! di = d_i
%! si = s_i
%! sj = s_j
%! F = \texttt{F}
%! T = \texttt{T}
%! forall = \forall
%! exists = \exists
%! empty = \emptyset
%! Sigma = \Sigma
%! eta = \eta
%! where = \textbf{where}
%! epsilon = \varepsilon
%! least = \phi
%! loop+ = loop_{+}
%! loop* = loop_{\times}
%! CatC = {\mathcal C}
%! CatA = {\mathcal A}
%! gamma = \gamma
%! {[ = \{
%! ]} = \}
%! elem = \in
%! dagger = ^\dagger
%! alpha = \alpha
%! beta = \beta
%! rho = \rho
%! @@ = \mu
%! @ = \,@\,
%! Pow = \mathcal{P}
%! Pi = \Pi
%! PiT = \Pi^{o}
%! PiEE = \Pi^{\eta\epsilon}_{+}
%! PiT = \Pi^{o}
%! PiTF = \Pi^{/}
%! bullet = \bullet
%! * = \times

\begin{document}
\conferenceinfo{POPL '14}{date, City.} 
\copyrightyear{2014}
\copyrightdata{[to be supplied]}

\title{\ldots}
\authorinfo{\ldots}
           {\ldots}
           {\ldots}

\maketitle

\begin{abstract}
\ldots
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background: {{Pi}} }
\label{sec:pi}

We review our language {{Pi}} providing the necessary background and
context for our higher-order extension.\footnote{The presentation in this
  section focuses on the simplest version of {{Pi}}. Other versions
  include the empty type, recursive types, and trace operators but these
  extensions are orthogonal to the higher-order extension emphasized in this
  paper.} The terms of {{Pi}} are not classical values and functions;
rather, the terms are isomorphism witnesses.  In other words, the terms of
{{Pi}} are proofs that certain ``shapes of values'' are isomorphic.
And, in classical Curry-Howard fashion, our operational semantics shows how
these proofs can be directly interpreted as actions on ordinary values which
effect this shape transformation. Of course, ``shapes of values'' are very
familiar already: they are usually called \emph{types}.  But frequently one
designs a type system as a method of classifying terms, with the eventual
purpose to show that certain properties of well-typed terms hold, such as
safety.  Our approach is different: we start from a type system, and then
present a term language which naturally inhabits these types, along with an
appropriate operational semantics.

\paragraph*{Data.}
We view {{Pi}} as having two levels:  it has traditional values, given by:
%subcode{bnf} include main
% values, v ::= () | left v | right v | (v, v)

\noindent and these are classified by ordinary types:
%subcode{bnf} include main
% value types, b ::= 1 | b + b | b * b

\noindent 
Types include the unit type {{1}}, sum types {{b1+b2}}, and product types
{{b1*b2}}.  Values include {{()}} which is the only value of type {{1}},
{{left v}} and {{right v}} which inject~{{v}} into a sum type, and
{{(v1,v2)}} which builds a value of product type.  These values should be
regarded as largely ancillary: we do not treat them as first-class citizens,
and they only occur when observing the effect of an isomorphism.

\paragraph*{Isomorphisms.} The important terms of {{Pi}} are witnesses to
(value) type isomorphisms.  They are also typed, by the shape of the (value)
type isomorphism they witness {{b <-> b}}.  Specifically, they are witnesses
to the following isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}r c l@{\hspace{-0.5pt}}l
%swap+ :&  b1 + b2 & <-> & b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) & <-> & (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b & <-> & b &: identr*
%swap* :&  b1 * b2 & <-> & b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) & <-> & (b1 * b2) * b3 &: assocr*
%dist :&~ (b1 + b2) * b3 & <-> & (b1 * b3) + (b2 * b3)~ &: factor

\noindent Each line of the above table introduces a pair of dual
constants\footnote{where {{swap*}} and {{swap+}} are self-dual.} that witness
the type isomorphism in the middle.  These are the base (non-reducible) terms
of the second, principal level of {{Pi}}. Note how the above has two
readings: first as a set of typing relations for a set of constants. Second,
if these axioms are seen as universally quantified, orientable statements,
they also induce transformations of the (traditional) values. The
(categorical) intuition here is that these axioms have computational content
because they witness isomorphisms rather than merely stating an extensional
equality.

The isomorphisms are extended to form a congruence relation by adding the
following constructors that witness equivalence and compatible closure:

%subcode{proof} include main
%@  ~
%@@ id : b <-> b 
%
%@ c : b1 <-> b2
%@@ sym c : b2 <-> b1
%
%@ c1 : b1 <-> b2
%@ c2 : b2 <-> b3
%@@ c1(;)c2 : b1 <-> b3
%---
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
%
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
\noindent The syntax is overloaded: we use the same symbol at the value-type level
and at the isomorphism-type level for denoting sums and products.  Hopefully
this will not cause undue confusion.

It is important to note that ``values'' and ``isomorphisms'' are completely
separate syntactic categories which do not intermix. The semantics of the
language come when these are made to interact at the ``top level'' via
\emph{application}: 
%subcode{bnf} include main
% top level term, l ::= c v

\noindent
To summarize, the syntax of {{Pi}} is given as follows.

\begin{definition}{(Syntax of {{Pi}})}
\label{def:Pi}
%subcode{bnf} include main
% value types, b ::= 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% iso.~types, t ::= b <-> b
% base iso ::= swap+ | assocl+ | assocr+ 
%     &|& unite | uniti | swap* | assocl* | assocr* 
%     &|& dist | factor 
% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
% top level term, l ::= c v
\end{definition}

The language presented above, at the type level, models a commutative ringoid
where the multiplicative structure forms a commutative monoid, but the
additive structure is just a commutative semigroup.  Note that the version of
{{Pi}} that includes the empty type with its usual laws exactly captures, at
the type level, the notion of a \emph{semiring} (occasionally called a
\emph{rig}) where we replace equality by isomorphism.  Semantically, {{Pi}}
models a \emph{bimonoidal category} whose simplest example is the category of
finite sets and bijections. In that interpretation, each value type denotes a
finite set of a size calculated by viewing the types as natural numbers and
each combinator {{c : b1 <-> b2}} denotes a bijection between the sets
denoted by~{{b1}} and~{{b2}}.

Operationally, the semantics of {{Pi}} is given using two mutually recursive
interpreters: one going forward and one going backwards. The use of {{sym}}
switches control from one evaluator to the other. We will present the
operational semantics in the next section along with the extension with
fractional types and values. For now, we state without proof that the
evaluation of well-typed combinators always terminates and that {{Pi}} is
logically reversible, i.e., that for all combinators {{c : b1 <-> b2}} and
values {{v1 : b1}} and {{v2 : b2}} we have the forward evaluation of {{c v1}}
produces {{v2}} iff the backwards evaluation of {{c v2}} produces {{v1}}.

As is usual with type theories, the coherence conditions (pentagonal and
triangle identity) are not explicit, but would correspond to certain identity
types being trivial.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Syntax of Fractionals}

The types of {{Pi}} correspond to the natural numbers. Our goal is to have a
type system that corresponds to the (positive) rational numbers.

\begin{definition}{(Syntax of {{PiTF}})}
\label{def:PiTF}
%subcode{bnf} include main
% value types, r ::= b / b
% values, f ::= (v |--> v amp c)
%
% iso.~types, t ::= r <=> r
% base iso ::= swap+ | assocl+ | assocr+ 
%     &|& unite | uniti | swap* | assocl* | assocr* 
%     &|& dist | factor 
%     &|& eta | epsilon
% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
% top level term, l ::= c v
\end{definition}

If we restrict our attention to types {{b / 1}} and values {{ () |--> v amp id}},
we recover {{Pi}}.

In algebra, the rational number {{a/b}} denotes 
$\{ (a','b) ~|~ a * b' = a' * b \}$. We essentially use the same denotation but
add computational content, i.e., the type {{a/b}} denotes
$\{ (a',b',c) ~|~ c : a * b' \leftrightarrow a' * b \}$ where $c$ is a {{Pi}}
combinator witnessing the equivalence of {{a * b'}} and {{a' * b}}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Higher-Order Reversible Functions}

To get from {{Pi}}~\cite{James:2012:IE:2103656.2103667} types and combinators
to a higher order setting, we hope to add \emph{fractional types} to the
language. Recall that to construct the (positive) rational numbers {{Q}}, we
can take pairs of positive integers {{(m, n)}} and quotient by the
equivalence relation that says {{(m1, n1)}} is equal to {{(m2, n2)}} if 
{{m1 * n2 = m2 * n1}}. We can write this as an inference rule as follows:

@@

%%%%%%%%%%%%%%%%%%%
\subsection{Defining trace}

@@

%%%%%%%%%%%%%%%%%%%
\subsection{Higher-Order Combinators}

@@

\bibliographystyle{abbrvnat}
\bibliography{cites}
\end{document}
